---
title: "CAGE Oikopleura (trans-splice start clustering)"
author: "Charles Plessy"
date: "28/07/2023"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load the CAGE data from BAM files.

See the [CAGEr_CTSS](./CAGEr_CTSS.Rmd) file for how the data was loaded.

```{r setup}
library("CAGEr")
library("ggplot2")
library("patchwork")
ce <- readRDS("ce.rds")

ce <- setColors(ce, c("red", "red", "red", "red",
                      "blue", "blue",
                      "pink", "pink", "pink", "pink",
                      "green", "green", "green", "green", "green", "green", "green", "green"))

gff <- rtracklayer::import("../AlignWithRNAseqPipelinePE/Bar2_p4.gm.gtf")
gff$type <- as.character(gff$type)
gff <- gff[gff$type %in% c("transcript", "intron", "exon")]
gff$gene_name <- gff$gene_id
```

# Process single-nucleotide-resolution signal

## Remove transcription start sites and TPM-normalise

The expression values SL peaks do not follow a power law distribution.  We can
see three components in the reverse cumulative distribution of the raw and
TPM-normalised expression values:

 - Less than 10 TPM;
 - between 10 and 1000 TPM;
 - more than 1000 TPM.

Also, the male sample has a different distribution of expression levels.

```{r}
ce_SL <- ce[,   ce$SLfound] |>
  resetCAGEexp() |>
  normalizeTagCount(method = "simpleTpm") |>
  annotateCTSS(gff)
p1_sl <- plotReverseCumulatives(ce_SL, fitInRange = NULL, values = "raw") +
  ggtitle("Before normalisation", sub="Trans-splicing sites") + theme_bw()
p2_sl <- plotReverseCumulatives(ce_SL, fitInRange = NULL, values = "normalized") +
  ggtitle("TPM normalised", sub="Trans-splicing sites") + theme_bw()
(p1_sl | p2_sl) + plot_layout(guides = 'collect')
```

## Define, label and study expression classes

First, we check that the same expression classes exist after adding the scores
from each library, and then we create a factor for these classes.

```{r}
CTSScoordinatesGRL <- function(object) {
  grl <- lapply(sampleLabels(object), CTSStagCountGR, object = object) |> GRangesList()
  names(grl) <- unname(sampleLabels(object))
  # Also pass the original colData?
  grl@metadata$colData <- colData(object)
  grl
}

# Explore properties of low-expressed CTSS
score(CTSScoordinatesGR(ce_SL)) <- CAGEr:::rowSums.RleDataFrame(CTSSnormalizedTpmDF(ce_SL)) / length(sampleLabels(ce_SL))
# Confirmation that the cutoffs are usable on pooled data.
CTSScoordinatesGR(ce_SL) |> plotReverseCumulatives(val = "norm", fit=NULL, group = "Pooled samples") +
  geom_vline(xintercept = c(10, 1000)) +
  ggtitle("Reverse-cumulative plot", subtitle = "Pooled samples (sum of scores)")
# Label CTSSes by expression classes
CTSScoordinatesGR(ce_SL)$CTSS_expr_range <- score(CTSScoordinatesGR(ce_SL)) |> decode() |> cut(c(0, 10, 1000, 1e6)) |> Rle()
```

The highly-expressed class is rare and close to promoters (actually, ATGs).  The mid-expressed
class contains a number of clusters that is roughly what we expect given than
~40% of the genes were reported to be trans-spliced.  The lowest tier of
expression comprises a very large number of sites but accounts for less than
1% of the library.

```{r}
# The highly expressed CTSS are rare but make most of the library.
tapply(CTSScoordinatesGR(ce_SL)$score, CTSScoordinatesGR(ce_SL)$CTSS_expr_range, length)
round(tapply(CTSScoordinatesGR(ce_SL)$score, CTSScoordinatesGR(ce_SL)$CTSS_expr_range, sum) / 1e6, 2)

plotAnnot_CTSS <- function(x, factor, gff, up = 100, down = 0, ...) {
  # plot percent of clusters, not percent of sum of expression of clusters
  x <- resetCAGEexp(x)
  x <- x[factor,]
  DF <- lapply(assay(x), \(y) Rle(as.numeric(y > 0))) |> DataFrame()
  assay(CTSStagCountSE(x)) <- DF
  x <- annotateCTSS(x, gff, up, down)
  x$librarySizes <- sapply(assay(x),sum)
  plotAnnot(x, "counts")
}

# A lot of the low-expression ones are still near promoters.
# Pay attention that here "promoters" are 100-bp windows strictly upstram
p_10_sl      <- plotAnnot_CTSS(ce_SL, (CTSScoordinatesGR(ce_SL)$CTSS_expr_range == "(0,10]" )       |> decode(), gff) + ggtitle("(0,10]")
p_1000_sl    <- plotAnnot_CTSS(ce_SL, (CTSScoordinatesGR(ce_SL)$CTSS_expr_range == "(10,1e+03]")    |> decode(), gff) + ggtitle("(10,1e+03]")
p_1000000_sl <- plotAnnot_CTSS(ce_SL, (CTSScoordinatesGR(ce_SL)$CTSS_expr_range == "(1e+03,1e+06]") |> decode(), gff) + ggtitle("(1e+03,1e+06]")
(p_10_sl / p_1000_sl / p_1000000_sl ) + plot_layout(guides = 'collect')
```

## AG dinucleotides at the level of CTSSSes

Non-AG dinucleotides are rare except in the low-score class.

```{r}
safePromoters <- function(gr, up, down) {
  p <- promoters(gr, up, down) |> suppressWarnings()
  p[width(p) == width(trim(p))]
}

AGplot <- function(gr) gr |> safePromoters(2,0) |> getSeq (x = BSgenome::getBSgenome('Bar2.p4')) |> table() |> prop.table() |> as.data.frame() |> ggplot() + aes(y = Var1, x = Freq) + geom_point() + scale_x_continuous("frequency", labels = scales::label_percent())
agplot1 <- AGplot(CTSScoordinatesGR(ce_SL)[CTSScoordinatesGR(ce_SL)$CTSS_expr_range == "(0,10]"] )
agplot2 <- AGplot(CTSScoordinatesGR(ce_SL)[CTSScoordinatesGR(ce_SL)$CTSS_expr_range == "(10,1e+03]"] )
agplot3 <- AGplot(CTSScoordinatesGR(ce_SL)[CTSScoordinatesGR(ce_SL)$CTSS_expr_range == "(1e+03,1e+06]"] )

(agplot1 / agplot2 / agplot3 ) + plot_layout(guides = 'collect')
```

# Sharp clustering of trans-splicing sites

## Paraclu parameters and run

We know from published results and from preliminary observations of our data
that the trans-splicing sites are single-base resolution and that sometimes
there are some secondary sites in the vicinity of a primary one.  Therefore we
will use the _paraclu_ clustering method, which allows us to cap the length of
the clusters to a maximum, which we will set to 10 nucleotides.  We also will
remove the CTSS with a expression level lower than 1 TPM in all samples.

Try different parameters

```{r try_paraclu_singleton_removal}
# Explore a subset of the data
ce_SL |> CTSSnormalizedTpmGR(1)|> head(10000) |>
  paraclu(maxLength = 10L, removeSingletons = TRUE, keepSingletonsAbove = 0) |> plotReverseCumulatives()
# Clustering does dot affect much the distribution.
ce_SL |> CTSSnormalizedTpmGR(1)|> head(10000) |>
  paraclu(maxLength = 10L, removeSingletons = TRUE, keepSingletonsAbove = 1) |> plotReverseCumulatives()
```

There is not much penalty keeping the peaks narrow.

```{r try_paraclu_maxlength}
ce_SL |> CTSSnormalizedTpmGR(1)|> head(1000) |> paraclu(maxLength = 10L, removeSingletons = TRUE, keepSingletonsAbove = 1) |> length()
ce_SL |> CTSSnormalizedTpmGR(1)|> head(1000) |> paraclu(maxLength = 20L, removeSingletons = TRUE, keepSingletonsAbove = 1) |> length()
ce_SL |> CTSSnormalizedTpmGR(1)|> head(1000) |> paraclu(maxLength = 50L, removeSingletons = TRUE, keepSingletonsAbove = 1) |> length()
ce_SL |> CTSSnormalizedTpmGR(1)|> head(1000) |> paraclu(maxLength = 200L, removeSingletons = TRUE, keepSingletonsAbove = 1) |> length()
ce_SL |> CTSSnormalizedTpmGR(1)|> head(1000) |> paraclu(maxLength = 500L, removeSingletons = TRUE, keepSingletonsAbove = 1) |> length()
```

```{r ce_SL_paraclu}
ce_SL <- ce_SL |> clusterCTSS( method = "paraclu"
                             , nrPassThreshold = 1 # Default.  We do not have replicates for all time points
                             , threshold = 1   # See above.  Note that it allows low-score CTSS supported in other samples.
                             , removeSingletons = TRUE
                             , keepSingletonsAbove = 1
                             , maxLength = 10L # Keep them sharp
                             , useMulticore = TRUE # Deigo
                             , nrCores = 8)    |>    # Yay !
                  cumulativeCTSSdistribution() |>
                  quantilePositions()
```

## QC of the Paraclu clusters

Most tag clusters are much narrower than the maximum width allowed.

```{r TC_IQ_width}
plotInterquantileWidth(ce_SL, clusters = "tagClusters", tpmThreshold = 3, qLow = 0.1, qUp = 0.9) + ggplot2::theme_bw()
```

Distribution of the expression values in CTSS and tag clusters are very similar,
except that the low-expressed ones were filtered out.

```{r}
tcs <- tagClustersGR(ce_SL)
tcs |> sapply(length)

p3_sl <- plotReverseCumulatives(tcs, val="norm", fit=NULL) + theme_bw() + ggtitle("Tag clusters", sub = "Trans-splicing sites")
(p2_sl + scale_y_log10(limits=c(1,1e5)) + ggtitle("CTSS", sub = "Trans-splicing sites") |
    p3_sl + scale_y_log10(limits=c(1,1e5))) + plot_layout(guides = 'collect')
```

```{r}
# Alternative representation as histograms
f <- function(name, tc) {
  tc[[name]] |> as.data.frame() |>
    dplyr::mutate(sampleName = name)
}
lapply(names(tcs), f, tcs) |> do.call(what = rbind) |> ggplot() +
  aes(x = score, fill = sampleName) +
  geom_histogram(bins = 100) +
  scale_x_log10() + facet_wrap(~sampleName) + theme_bw()
```

```{r}
TCs <- tagClustersGR(ce_SL)
TCs_dom <-endoapply(TCs, \(gr) {
  gr$dominant_ctss$expr_range <- score(gr) |> decode() |> cut(c(0, 10, 1000, 1e6)) |> Rle()
  gr$dominant_ctss
})
distanceToATGbyScore <- function(TCs_dom, gff, expr_range) {
  lapply(TCs_dom, \(gr) {
             hits <- distanceToNearest(gr[gr$expr_range == expr_range],
                                       promoters(gff[gff$type == "transcript"], 0, 1))
             mcols(hits)$distance
  }) |> sapply(summary)
}
distanceToATGbyScore(TCs_dom, gff, "(0,10]")
distanceToATGbyScore(TCs_dom, gff, "(10,1e+03]")
distanceToATGbyScore(TCs_dom, gff, "(1e+03,1e+06]")
```

# Aggregate the CAGE tags

```{r}
# First I ran with a threshold of 1, but it captured a lot of ectopic trans-splicing
# at splice junctions, which we do not want to use for updating annotation and
# defining operons.
ce_SL <- ce_SL |> 
  aggregateTagClusters(maxDist = 10L, tpmThreshold = 10, excludeSignalBelowThreshold = FALSE) |>  # See also the score distribution
  cumulativeCTSSdistribution(clusters = "consensusClusters") |>
  quantilePositions(clusters = "consensusClusters") |>
  annotateConsensusClusters(gff, up = 100, down = 0)
```

The clusters for the trans-spliced reads are short and contain most of the reads.

```{r}
x <- consensusClustersGR(ce_SL)
(100 - ce_SL$outOfClusters / ce_SL$librarySizes) |> round(3)
plotReverseCumulatives(x)
# No score is lower than our 10 TPM threshold
hist(log10(decode(score(x))), br = 100)
x$maxTPM <- ce_SL |> consensusClustersSE() |> assay("normalized") |> rowMax() |> unname() |> Rle()
hist(log10(decode(x$maxTPM)), br = 100)
consensusClustersGRL <- function(object) {
  ccGRreallyGR <- function(object, label) {
    gr <- GRanges(consensusClustersGR(object, label))
    score(gr) <- Rle(score(gr))
    gr
  }
  grl <- lapply(sampleLabels(object), ccGRreallyGR, object = object) |> GRangesList()
  names(grl) <- unname(sampleLabels(object))
  # Also pass the original colData?
  grl@metadata$colData <- colData(object)
  grl
}
clustersGRL <- consensusClustersGRL(ce_SL) |>
  endoapply(\(gr) {
    #gr$expr_range <- score(gr) |> decode() |> cut(c(0, 10, 1000, 1e6), include.lowest = TRUE)
    gr$expr_range <- score(gr) |> decode() |> cut(c(0, 10, 1000, 1e6), include.lowest = FALSE)
    gr$expr_range <- gr$expr_range |> factor(levels = append("[0]", levels(gr$expr_range))) 
    gr$expr_range[is.na(gr$expr_range)] <- "[0]"
    gr
  })
clustersGRL |> sapply(\(gr) table(gr$expr_range))

filterGRL <- function(grl, x) {
  endoapply(grl, \(gr) {
    gr[gr$expr_range == x]
  })
}
myPlotAnnot <- function(grl, range) {
  grl |> filterGRL(range) |> plotAnnot("counts", range)
}

p1_cc <- myPlotAnnot(clustersGRL, "[0,10]")
p2_cc <- myPlotAnnot(clustersGRL, "(10,1e+03]")
p3_cc <- myPlotAnnot(clustersGRL, "(1e+03,1e+06]")

(p1_cc / p2_cc / p3_cc) + plot_layout(guides = 'collect')


plotReverseCumulatives(clustersGRL)
plotAnnot(clustersGRL, "counts", "Consensus clusters")
x$expr_range <- x$maxTPM |> decode() |> cut(c(0, 10, 1000, 1e6))
table(x$expr_range)
plotAnnot(clustersGRL, "counts", "Consensus clusters", facet="expr_range")


# A lot of the low-expression ones are still near promoters.
# Pay attention that here "promoters" are 200-bp windows.
p_10 <- plotAnnot_CTSS(x[x$expr_range == "(0,10]", ] , gff) + ggtitle("(0,10]")
p_1000 <- plotAnnot_CTSS(ce_SL[CTSScoordinatesGR(ce_SL)$CTSS_expr_range == "(10,1e+03]", ], gff) + ggtitle("(10,1e+03]")
p_1000000 <- plotAnnot_CTSS(ce_SL[CTSScoordinatesGR(ce_SL)$CTSS_expr_range == "(1e+03,1e+06]", ], gff) + ggtitle("(1e+03,1e+06]")
(p_10 / p_1000 / p_1000000 ) + plot_layout(guides = 'collect')


consensusClustersGR(ce_SL, 1) |> score() |> decode() |> log10() |> hist(100)
consensusClustersGR(ce_SL, 3) |> score() |> decode() |> log10() |> hist(100)
plotInterquantileWidth(ce_SL, clusters = "consensusClusters", tpmThreshold = 3, qLow = 0.1, qUp = 0.9)

rowRanges(consensusClustersSE(ce_SL))$normExprTot <- rowSums(consensusClustersSE(ce_SL)|>assay("normalized")) 
rowRanges(consensusClustersSE(ce_SL))$normExprRange <- cut(rowRanges(consensusClustersSE(ce_SL))$normExprTot, c(0, 10, 1000, 1e6))
rowRanges(consensusClustersSE(ce_SL))$normExprRange  |> summary()

```

# Export Consensus clusters

## Trans-splicing sites

Paint in grey those that do not have `AG` upstream.

```{r}
cctrack_SL <- exportToTrack(ce_SL, "consensusClusters", qLow = 0.1, qUp = 0.9)
cctrack_SL@trackLine@description <- "CAGE Consensus Clusters for trans splicing sites"
cctrack_SL@trackLine@name <- "SL"
# Flat AG
cctrack_SL$itemRgb <- ifelse(flagByUpstreamSequences(rowRanges(consensusClustersSE(ce_SL))$dominant_ctss, "AG"), "black", "grey")

cctrack_SL[cctrack_SL$itemRgb == "black"] |> score() |> decode() |> log10() |> hist(br=100)
cctrack_SL[cctrack_SL$itemRgb == "grey"]  |> score() |> decode() |> log10() |> hist(br=100)

rtracklayer::export.bed(cctrack_SL, "clusters_SL.bed")
```

```{r}
cctrack_no <- exportToTrack(ce_no, "consensusClusters", qLow = 0.1, qUp = 0.9)
cctrack_no@trackLine@description <- "CAGE Consensus Clusters for transcription start sites"
cctrack_no@trackLine@name <- "TSS"
# Flat the clusters of width 1.
cctrack_no$itemRgb <- ifelse(width(cctrack_no) > 1, "black", "grey")

rtracklayer::export.bed(cctrack_no, "clusters_no.bed")
```

```
zenbu_upload -url https://fantom.gsc.riken.jp/zenbu/ -file clusters_SL.bed -name ConsensusClustersSL -desc 'Consensus Clusters for trans-splicing CAGE data. Uploaded on 2023012701.' -assembly Oidioi_Bar2_p4.Flye -collab orXQCELWOZfm7KLNMMpPdD
zenbu_upload -url https://fantom.gsc.riken.jp/zenbu/ -file clusters_no.bed -name ConsensusClustersNoSL -desc 'Consensus Clusters for transription start CAGE data. Uploaded on 2023012701.' -assembly Oidioi_Bar2_p4.Flye -collab orXQCELWOZfm7KLNMMpPdD
```

# Export the _CAGEexp_ object.

```{r}
saveRDS(ce_SL, "ce_SL.rds")
```

# AG motifs

```{r AG_motifs}

TSSlogo(cctrack_SL[cctrack_SL$itemRgb == "black"], 40, 20)
TSSlogo(cctrack_SL[cctrack_SL$itemRgb == "black"], 16, 16) # Very similar to Danks 2015 !!
TSSlogo(cctrack_SL[cctrack_SL$itemRgb == "grey"], 40, 20)
TSSlogo(cctrack_SL[cctrack_SL$itemRgb == "grey"], 16, 16)

seqLogo_GR(cctrack_no, 35, 15)

cctrack_SL$dominant_ctss |> promoters(2,0) |> getSeq (x = BSgenome::getBSgenome('Bar2.p4')) |> table() |> prop.table() |> tibble::as_tibble() |> ggplot() + aes(y = upstreamSeq, x = n) + geom_point() + scale_x_continuous("frequency", labels = scales::label_percent())
```


# Hanabi !

```{r hanabi_plots}
grp <- c("Female", "Female",
                      "Male",
                      "Juvenile", "Juvenile",
                      "Embryo", "Embryo", "Embryo", "Embryo")

hanabi(CTSStagCountDF(ce_SL)) |>                hanabiPlot(group = grp, col = c("red", "blue", "pink", "green"))
hanabi(consensusClustersSE(ce_SL)|> assay()) |> hanabiPlot(group = grp, col = c("red", "blue", "pink", "green"), legend = "bottomright")
```

# Session information

```{r sessionInfo}
sessionInfo()
```